---
title: "explore_demographics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE, include = FALSE}
# RMarkdown's insistence on file paths and packages in the same file
library(tidyverse) 
projectFolder  <- "~/github/fisher_mental_health"
data <- file.path(projectFolder, "data/csv")

```


Bring in demographic files

```{r}

demog <- read.csv(file.path(data, "tbl_A_SocioDemog.csv"), na.strings = c("", "NA"))

```

According to the questionnaire, this file pertains to the mother only. However, there are 4 males listed in the file

```{r}
table(demog$A1_Gender)
summary(demog$A1_Gender)

head(demog %>% 
       select(ParticipantID, A1_Gender) %>% 
       filter(A1_Gender == "1 - Male"))
```

Investigating further: Of the ParticipantIDs that are associated with males, do they appear as females at a different time point?

```{r}
subset <- demog %>% select(TimePoint, ParticipantID, A1_Gender)
head(subset %>% filter(ParticipantID == "E0512"))  # this is a woman
head(subset %>% filter(ParticipantID == "S0053"))  # this is unclear
head(subset %>% filter(ParticipantID == "S0103"))  # this is unclear
head(subset %>% filter(ParticipantID == "W0072"))  # this is a woman
rm(subset)
```

As a separate check, I am going to look at the fisher access survey which is only asked of fishers, all of whom are male. 

```{r}
a <- read.csv(file.path(data, "tbl_I_FisheryAccess.csv"), na.strings=c("", "NA"))

# create a vector of fisher IDs. Access qx only asked of fishers, so pulling from there.
fishers_only <- a %>% 
  select(ParticipantID) %>% 
  unique() %>% 
  mutate(male = 1)

dim(fishers_only)      # 269 unique fishers in the sample

head(fishers_only %>% filter(ParticipantID == "E0512"))  # this is a woman
head(fishers_only %>% filter(ParticipantID == "S0053"))  # this is confirmed to be a man
head(fishers_only %>% filter(ParticipantID == "S0103"))  # this is confirmed to be a man
head(fishers_only %>% filter(ParticipantID == "W0072"))  # this is a woman

rm(a) # removing a, but keeping fishers_only for reference
```


Participant

To complement `fishers_only` file, I will generate a `mothers_only` file containing ParticipantIDs of mothers and excluding the two ParticipantIDs above that have been confirmed as male.

```{r}

mothers_only <- demog %>% 
  filter(ParticipantID != "S0053" & ParticipantID != "S0103") %>% 
  select(ParticipantID) %>% 
  unique() %>% 
  mutate(female = 1)

dim(mothers_only)      #313 unique mothers in the sample

```

ParticipantID E0092 is listed as a fisher and a mother, but is a mother, so removing her from fishers_only. 
ParticipantID W0233 is listed as a fisher and a mother, but is a fisher, so removing him from mothers_only

```{r}

fishers_only <- fishers_only %>% filter(ParticipantID != "E0092")
mothers_only <- mothers_only %>% filter(ParticipantID != "W0233")

```



Saving the workspace so that I can knit in subsequent Rmd documents
```{r}
save.image("scripts/demog")
```

