---
title: "Initial Data Exploration"
author: "Liz Bageant"
date: "12/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

This is the first initial exploration of the Mfangano data for the fisher mental health paper. Will keep notes simple for now. 

```{r , echo = FALSE, include = FALSE}
#install.packages("tidyverse")
#install.packages("skimr")
#install.packages("ggpubr")
library(ggpubr) 
library(tidyverse) 
library(skimr)
library(infer) #install.packages("infer")

```

Make folder paths

```{r}
projectFolder  <- "~/github/fisher_mental_health"
data <- file.path(projectFolder, "data/csv")
```



```{r}
access <- read.csv(file.path(data, "tbl_I_FisheryAccess.csv"), na.strings=c("", "NA"))
morbidity <- read.csv(file.path(data, "tbl_L_Morbidity.csv"))

# create a vector of fisher IDs. Access qx only asked of fishers, so pulling from there.
fishers_only <- access %>% 
  select(ParticipantID) %>% 
  unique()

dim(fishers_only) # 269 unique fishers in the sample (though not all of them fished at all time periods)

```








**COMBINE MORBIDITY AND FISHER BEHAVIOR DATA FRAMES**
```{r}

fishers <- full_join(fisher_mh_subscale, fishing_behavior, key = c("ParticipantID", "TimePoint"))

# cleaning up workspace before visualizing some things

rm(access, fisher_mh, fisher_mh_recode, fishing_behavior, fishing_behavior_f, fishing_behavior_nf, maxx, minn, mh_sub, morbidity, x, moshiv)

# using TimePoint to create "period" variable that we can treat as numeric for time series analysis
table(fishers$TimePoint, useNA = "always")

fishers <- fishers %>% 
  mutate(period = as.numeric(recode(TimePoint,
                         "0" = "1",
                         "3" = "2",
                         "6" = "3", 
                         "12" = "4",
                         "18" = "5",
                         "24" = "6",
                         .default = "99"))) # i think this sets missing values to 99

table(fishers$period)


```






**VISUALIZE FISHER MENTAL HEALTH DATA**

Key takeaways from plots/tables below, as of 12/24/2021:

* The variation is not impressive. 
* The bulk of the sample is in the "some of the time" or "none of the time" categories, not "most of the time" or "all of the time". 
* The bulk of the movement seems to be between "some of the time" and "none of the time" too

Ideas for next steps: 

* What do mothers' mental health subscales look like? 
  * How different are they from fishers?
  * Are they collected at every time point?



Plotting fisher mental health subscale over time.

```{r}

ggplot(data = fishers) +
  geom_jitter(mapping = aes(x = period , y = mh_subscale)) 
ggsave("../images/mh_scatter_by_period.png")

histogram <- ggplot(fishers) +
  geom_histogram(mapping = aes(x = mh_subscale)) +
  facet_wrap(~ period)
ggsave("../images/mh_histogram_by_period.png")

ggplot(fishers) +
  geom_density(mapping = aes(x = mh_subscale)) +
  facet_wrap(~ period)
ggsave("../images/mh_density_by_period.png")

```


I now want to know some summary statistics. What is the overall mean, the mean by period and the within-person mean. 
```{r, results = "hold"}

# summary statistics by period
fishers %>% 
  group_by(period) %>% 
  summarise(mean = mean(mh_subscale), median = median(mh_subscale), sd = sd(mh_subscale)) %>% 
  knitr::kable(caption = "Fisher MH subscale by period")

ggplot(fishers, aes(x = as.factor(period), y = mh_subscale)) +
  geom_boxplot() +
  labs(title = "Mental Health Subscale by period")
ggsave("../images/mh_boxplot_by_period.png")

```

Quick calculation of within-household mean and SD. 
```{r, results = "hold"}
within_means <- fishers %>% 
  group_by(ParticipantID) %>% 
  summarise(mean = mean(mh_subscale), sd = sd(mh_subscale)) 
within_means %>% 
  summarise(mean1 = mean(mean), sd = sd(mean)) %>% 
  knitr::kable(caption = "Within-fisher mean and SD")
            
```

```{r}

ggplot(within_means) +
  geom_histogram(mapping = aes(x = mean)) +
  labs(title = "Distribution of within-household means of Mental Health Subscale scores")

ggplot(within_means) +
  geom_histogram(mapping = aes(x = sd)) +
  labs(title = "Distribution of within-household standard deviations of Mental Health Subscale scores")

```


**VISUALIZE FISHER BEHAVIOR**

Fishers' decision to fish over time is pretty stable on average across fishers. On average, 60-70% of fishers fished in the past 3 months.
```{r}

# Examining missing values of fisher behavior. There are 15 cases with no clear pattern in terms of time point. Individual EOO92 is missing five time points.
fishers %>% 
  filter(is.na(fished_i1)) # all missing cases of fished_i1 --> completely missing fisher behavior
fishers %>% 
  filter(is.na(period))

# Filtering out missing fisher behavior data and summarizing by period
fishers %>% 
  filter(!is.na(fished_i1)) %>% 
  group_by(period) %>% 
  summarise(mean = mean(fished_i1, na.rm = TRUE), sd = sd(fished_i1, na.rm = TRUE)) %>% 
  ggplot(decision, mapping = aes(x = period, y = mean)) +
  geom_path(group = 1) +
  ylim(0, 1) +
  labs(title = "Share of fishers who fished in any given period", y = "Share", x = "Period")
  

```



Fishing behavior among those that fished

Among those that fished, there is a lot of variation in **perceptions of environmental changes** that affect fishing. 

```{r, results = "hold"}
print("In the last 3 months has there been an effect on the following: Environmental changes that affect fishing (examples: algae, water hyacinth)")
table(fishers$change_enviro_i310a, fishers$period, dnn=c("Yes/No", "Period"))
```

Variation in eports of **weather changes** affecting ability to fish

```{r}
print("Have any of the following shocks affected your ability to fish in the last 3 months? Weather changes")
table(fishers$shock_weather_i311, fishers$TimePoint, dnn=c("Weather Changes", "Period"))
```

Limited reports of **water hyacinth**
```{r}
table(fishers$shock_hyacinth_i311, fishers$TimePoint, dnn=c("Hyacinth", "Period"))
```

Almost everyone reported **severe fish decline** to the point where 
```{r}
table(fishers$shock_fishdecline_i311, fishers$TimePoint, dnn=c("Severe fish decline", "Period"))
```



Fishing behavior among those that did not fish
Among those that did not fish, very few cited **environmental changes that contributed to them not fishing**

```{r, results = "hold"}

table(fishers$stop_decision_enviro_i245a, fishers$TimePoint, dnn=c("Yes/No", "Period"))

```

Did whether or not you are allowed to fish (regulations) affect your decision to stop fishing?

```{r}
table(fishers$stop_decision_regulation_i244a, fishers$TimePoint, dnn=c("Yes/No", "Period"))
```




