---
title: "explore_access"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Bring in fish access file

```{r}
access <- read.csv(file.path(data, "tbl_I_FisheryAccess.csv"), na.strings=c("", "NA"))

# create a vector of fisher IDs. Access qx only asked of fishers, so pulling from there.
fishers_only <- access %>% 
  select(ParticipantID) %>% 
  unique()

dim(fishers_only) # 269 unique fishers in the sample (though not all of them fished at all time periods)

```


**ORGANIZE FISHING ACCESS DATA**

There is a skip pattern in the questionnaire. 
"Access Limitations" questions differ slightly between fisher and non-fishers 
depending on whether they answer yes or no to I1: *"Have you fished or hired labor to fish in the last 3 months"*. 
If "no", answer questions in I2. If yes, answer questions in I3. 


Create a dataset for fishing behavior and motivation for **those who did not fish** in the past 3 months

Note: Variable names have suffix that represents their question number in first follow-up survey located at: Survey Instruments and Instructions/Questionnaires-03 Month Follow-Up. 
e.g. Module I, question 3.1 --> i31. 

```{r}
fishing_behavior_nf <- access %>% 
  mutate(fished_or_hired_3months_i1 = as.numeric(I1_FishedOrHiredLaborLastThreeMosYN),
         hired_ever_i21 = as.numeric(I21_HireLabor),
         leased_ever_i21 = as.numeric(I21_GoFishingUnderLease),
         self_ever_i21 = as.numeric(I21_GoFishingSelfEmployed),
         forpay_ever_i21 = as.numeric(I21_WorkAsHiredLabor),
         last_fished_i22 = as.character(I22_LastTimeFished), 
         stop_reason_i231 = as.character(I231_ReasonsStoppedFishing), 
         stop_reason_important_i232 = as.character(I232_MostImportantReasonStoppedFishing),
         stop_decision_money_i241a = as.numeric(I241_MoneyReceivedYN),
         stop_decision_money_why_i241b = as.character(I241_MoneyReceivedHow), 
         stop_decision_time_i242a = as.numeric(I242_TimeSpentFishingYN),
         stop_decision_time_why_i242b = as.character(I242_TimeSpentFishingHow),
         stop_decision_difficulty_i243a = as.numeric(I243_DifficultyFishingYN), 
         stop_decision_difficulty_why_i243b = as.character(I243_DifficultyFishingHow), 
         stop_decision_regulation_i244a = as.numeric(I244_RegulationsYN),
         stop_decison_regulation_why_i244b = as.character(I244_RegulationsHow), 
         stop_decision_enviro_i245a = as.numeric(I245_EnvironChangesYN), 
         stop_decision_enviro_why_i245b = as.character(I245_EnvironChangesHow)) %>% 
  select(TimePoint, ParticipantID, fished_or_hired_3months_i1, hired_ever_i21, 
         self_ever_i21, leased_ever_i21, forpay_ever_i21, last_fished_i22, starts_with("stop_"))
```

The above data frame should contain information for only those that answered "Yes" 
though there are some 0s that should be missing.

```{r}
colSums(!is.na(fishing_behavior_nf)) ## counting non-missing values
colSums(is.na(fishing_behavior_nf)) ## counting missing values
```

Replacing values of non-fisher variables with NA for those who fished (fishers = `fished_or_hired_3months == 1`)

```{r}

fishing_behavior_nf %>% 
  mutate_at(vars(hired_ever_i21, self_ever_i21, forpay_ever_i21, last_fished_i22,
                 starts_with("stop_")), ~replace(.,fished_or_hired_3months_i1 == 1, NA))
```

Create a dataset for fishing behavior and motivation for **those who reported fishing** in the past 3 months

```{r}
fishing_behavior_f <- access %>% 
  mutate(fished_or_hired_3months_i1 = as.numeric(I1_FishedOrHiredLaborLastThreeMosYN),
         hired_now_i31 = as.numeric(I31_HireLabor), 
         leased_now_i31 = as.numeric(I31_GoFishingUnderLease),
         self_now_i31 = as.numeric(I31_GoFishingSelfEmployed),
         forpay_now_i31 = as.numeric(I31_WorkAsHiredLabor),
         enough_fish_i32 = as.character(I32_LastThreeMonthsEnufFish), 
         quantity_change_i331 = as.character(I331_LastThreeMonthsQtyofFishChanged),
         quantity_effect_frequency_i332 = as.numeric(I332_ChangeAffectedHowOftenFishYN),
         quantity_effect_frequency_why_i333 = as.character(I333_ChangeAffectedHowOftenFishHow),
         quantity_effect_investment_i334 = as.numeric(I334_ChangeAffectedHowMuchMoneySpendYN),
         quantity_effect_investment_why_i335 = as.character(I335_ChangeAffectedHowMuchMoneySpendHow),
         limits_use_i34 = as.character(I34_LastThreeMonthsLimitsAbilityToUseFish),
         limits_use_most_i35 = as.character(I35_LastThreeMonthsMostImportantLimit),
         change_money_i36a = as.numeric(I36a_MoneyCanMakeYN),
         change_money_cause_i36b = as.character(I36b_MoneyCanMakeCauses),
         change_money_direction_i36c = as.character(I36c_MoneyCanMakeIncreaseDecrease),
         change_time_i37a = as.numeric(I37a_TimeSpentFishingYN),
         change_time_cause_i37b = as.character(I37b_TimeSpentFishingCauses),
         change_time_direction_i37c = as.character(I37c_TimeSpentFishingIncreaseDecrease),
         change_difficulty_i38a = as.numeric(I38a_DifficultyFishingYN),
         change_difficulty_cause_i38b = as.character(I38b_DifficultyFishingCauses),
         change_difficulty_direction_i38c = as.character(I38c_DifficultyFishingIncreaseDecrease),
         change_regulation_i39a = as.numeric(I39a_AllowedToFishYN),
         change_regulation_cause_i39b = as.character(I39b_AllowedToFishCauses),
         change_regulation_direction_i39c = as.character(I39c_AllowedToFishIncreaseDecrease),
         change_enviro_i310a = as.numeric(I310a_EnvironChangesYN),
         change_enviro_cause_i310b = as.character(I310b_EnvironChangesCauses),
         change_enviro_direction_i310c = as.character(I310c_EnvironChangesIncreaseDecrease), 
         shock_gearstolen_i311 = as.numeric(I311_GearsStolenYN),
         shock_gearconfiscated_i311 = as.numeric(I311_GearsConfiscatedYN),
         shock_hyacinth_i311 = as.numeric(I311_WaterHyacinthYN),
         shock_weather_i311 = as.numeric(I311_WeatherChangesYN),
         shock_fishdecline_i311 = as.numeric(I311_SevereFishDeclineYN),
         shock_pricedecline_i311 = as.numeric(I311_SevereFishPriceDeclineYN),
         shock_inputprice_i311 = as.numeric(I311_IncreaseGearLaborPricesYN),
         shock_illness_i311 = as.numeric(I311_IllnessFamilyMemberYN),
         shock_death_i311 = as.numeric(I311_DeathFamilyMemberYN),
         shock_other_i311 = as.numeric(I311_OtherYN),
         shock_otherspecify_i311 = as.character(I311_OtherSpecify)) %>% 
  select(TimePoint, ParticipantID, fished_or_hired_3months_i1, leased_now_i31, 
         self_now_i31, hired_now_i31, forpay_now_i31, enough_fish_i32, starts_with(c("quantity", "limits", "change", "shock")))
```

Replacing values of fisher variables with NA for those who didn't fish (non-fishers = `fished_or_hired_3months == 0`)

```{r}

fishing_behavior_f <- fishing_behavior_f %>% 
  mutate_at(vars(hired_now_i31, self_now_i31, forpay_now_i31, leased_now_i31, 
                 starts_with(c("quantity_", "limits", "change", "shock"))), ~replace(.,fished_or_hired_3months_i1 == 0, NA))
            
```


In some time points, the fishing access questionnaire includes likert scale questions about perecpetions/preferences related to fishing. Currently excluding these because it's not clear how I will use them, but I may bring them in later.

Next step is to combine the fisher and non-fisher data frames. The fisher and non-fisher data frames contain some overlapping variables (e.g. QI.1) but most variables are specific to whether an individual fished or did not fish.
```{r}
fishing_behavior <- full_join(fishing_behavior_f, fishing_behavior_nf, by = c("ParticipantID", "TimePoint"))

# cleaning up where the join duplicated fished_or_hired_3months_i1

x <- fishing_behavior %>% 
  mutate(diff = fished_or_hired_3months_i1.x - fished_or_hired_3months_i1.y) %>% 
  select(fished_or_hired_3months_i1.x, fished_or_hired_3months_i1.y, diff) 
table(x$diff)

fishing_behavior <- fishing_behavior %>% 
  mutate(fished_i1 = fished_or_hired_3months_i1.x) 

# What is frequency of fishing vs not fishing, among fishers?
table(fishing_behavior$fished_i1)

```


