---
title: "Initial Data Exploration"
author: "Liz Bageant"
date: "12/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "hide")
```

This is the first initial exploration of the Mfangano data for the fisher mental health paper. Will keep notes simple for now. 

```{r , echo = FALSE, include = FALSE}
#install.packages("tidyverse")
#install.packages("skimr")

library(tidyverse) 
library(skimr)

```

Make folder paths

```{r}
projectFolder  <- "~/github/fisher_mental_health"
data <- file.path(projectFolder, "data/csv")
```

Bring in fish access file and morbidity fileS

```{r}
access <- read.csv(file.path(data, "tbl_I_FisheryAccess.csv"), na.strings=c("", "NA"))
morbidity <- read.csv(file.path(data, "tbl_L_Morbidity.csv"))
```

**ORGANIZE AND EXPLORE FISHING ACCESS QUESTIONS**

There is a skip pattern in the questionnaire. 
"Access Limitations" questions differ slightly between fisher and non-fishers 
depending on whether they answer yes or no to I1: *"Have you fished or hired labor to fish in the last 3 months"*. 
If "no", answer questions in I2. If yes, answer questions in I3. 


Create a dataset for fishing behavior and motivation for **those who did not fish** in the past 3 months

Note: Variable names have suffix that represents their question number in first follow-up survey located at: Survey Instruments and Instructions/Questionnaires-03 Month Follow-Up. 
e.g. Module I, question 3.1 --> i31. 

```{r}
fishing_behavior_nf <- access %>% 
  mutate(fished_or_hired_3months_i1 = as.numeric(I1_FishedOrHiredLaborLastThreeMosYN),
         hired_ever_i21 = as.numeric(I21_HireLabor),
         leased_ever_i21 = as.numeric(I21_GoFishingUnderLease),
         self_ever_i21 = as.numeric(I21_GoFishingSelfEmployed),
         forpay_ever_i21 = as.numeric(I21_WorkAsHiredLabor),
         last_fished_i22 = as.character(I22_LastTimeFished), 
         stop_reason_i231 = as.character(I231_ReasonsStoppedFishing), 
         stop_reason_important_i232 = as.character(I232_MostImportantReasonStoppedFishing),
         stop_decision_money_i241a = as.numeric(I241_MoneyReceivedYN),
         stop_decision_money_why_i241b = as.character(I241_MoneyReceivedHow), 
         stop_decision_time_i242a = as.numeric(I242_TimeSpentFishingYN),
         stop_decision_time_why_i242b = as.character(I242_TimeSpentFishingHow),
         stop_decision_difficulty_i243a = as.numeric(I243_DifficultyFishingYN), 
         stop_decision_difficulty_why_i243b = as.character(I243_DifficultyFishingHow), 
         stop_decision_regulation_i244a = as.numeric(I244_RegulationsYN),
         stop_decison_regulation_why_i244b = as.character(I244_RegulationsHow), 
         stop_decision_enviro_i245a = as.numeric(I245_EnvironChangesYN), 
         stop_decision_enviro_why_i245b = as.character(I245_EnvironChangesHow)) %>% 
  select(TimePoint, ParticipantID, fished_or_hired_3months_i1, hired_ever_i21, 
         self_ever_i21, leased_ever_i21, forpay_ever_i21, last_fished_i22, starts_with("stop_"))
```

The above data frame should contain information for only those that answered "Yes" 
though there are some 0s that should be missing.

```{r}
colSums(!is.na(fishing_behavior_nf)) ## counting non-missing values
colSums(is.na(fishing_behavior_nf)) ## counting missing values
```

Replacing values of non-fisher variables with NA for fishers (fishers = `fished_or_hired_3months == 1`)

```{r}

fishing_behavior_nf %>% 
  mutate_at(vars(hired_ever_i21, self_ever_i21, forpay_ever_i21, last_fished_i22,
                 starts_with("stop_")), ~replace(.,fished_or_hired_3months_i1 == 1, NA))
```

Create a dataset for fishing behavior and motivation for **those who reported fishing** in the past 3 months

```{r}
fishing_behavior_f <- access %>% 
  mutate(fished_or_hired_3months_i1 = as.numeric(I1_FishedOrHiredLaborLastThreeMosYN),
         hired_now_i31 = as.numeric(I31_HireLabor), 
         leased_now_i31 = as.numeric(I31_GoFishingUnderLease),
         self_now_i31 = as.numeric(I31_GoFishingSelfEmployed),
         forpay_now_i31 = as.numeric(I31_WorkAsHiredLabor),
         enough_fish_i32 = as.character(I32_LastThreeMonthsEnufFish), 
         quantity_change_i331 = as.character(I331_LastThreeMonthsQtyofFishChanged),
         quantity_effect_frequency_i332 = as.numeric(I332_ChangeAffectedHowOftenFishYN),
         quantity_effect_frequency_why_i333 = as.character(I333_ChangeAffectedHowOftenFishHow),
         quantity_effect_investment_i334 = as.numeric(I334_ChangeAffectedHowMuchMoneySpendYN),
         quantity_effect_investment_why_i335 = as.character(I335_ChangeAffectedHowMuchMoneySpendHow),
         limits_use_i34 = as.character(I34_LastThreeMonthsLimitsAbilityToUseFish),
         limits_use_most_i35 = as.character(I35_LastThreeMonthsMostImportantLimit),
         change_money_i36a = as.numeric(I36a_MoneyCanMakeYN),
         change_money_cause_i36b = as.character(I36b_MoneyCanMakeCauses),
         change_money_direction_i36c = as.character(I36c_MoneyCanMakeIncreaseDecrease),
         change_time_i37a = as.numeric(I37a_TimeSpentFishingYN),
         change_time_cause_i37b = as.character(I37b_TimeSpentFishingCauses),
         change_time_direction_i37c = as.character(I37c_TimeSpentFishingIncreaseDecrease),
         change_difficulty_i38a = as.numeric(I38a_DifficultyFishingYN),
         change_difficulty_cause_i38b = as.character(I38b_DifficultyFishingCauses),
         change_difficulty_direction_i38c = as.character(I38c_DifficultyFishingIncreaseDecrease),
         change_regulation_i39a = as.numeric(I39a_AllowedToFishYN),
         change_regulation_cause_i39b = as.character(I39b_AllowedToFishCauses),
         change_regulation_direction_i39c = as.character(I39c_AllowedToFishIncreaseDecrease),
         change_enviro_i310a = as.numeric(I310a_EnvironChangesYN),
         change_enviro_cause_i310b = as.character(I310b_EnvironChangesCauses),
         change_enviro_direction_i310c = as.character(I310c_EnvironChangesIncreaseDecrease), 
         shock_gearstolen_i311 = as.numeric(I311_GearsStolenYN),
         shock_gearconfiscated_i311 = as.numeric(I311_GearsConfiscatedYN),
         shock_hyacinth_i311 = as.numeric(I311_WaterHyacinthYN),
         shock_weather_i311 = as.numeric(I311_WeatherChangesYN),
         shock_fishdecline_i311 = as.numeric(I311_SevereFishDeclineYN),
         shock_pricedecline_i311 = as.numeric(I311_SevereFishPriceDeclineYN),
         shock_inputprice_i311 = as.numeric(I311_IncreaseGearLaborPricesYN),
         shock_illness_i311 = as.numeric(I311_IllnessFamilyMemberYN),
         shock_death_i311 = as.numeric(I311_DeathFamilyMemberYN),
         shock_other_i311 = as.numeric(I311_OtherYN),
         shock_otherspecify_i311 = as.character(I311_OtherSpecify)) %>% 
  select(TimePoint, ParticipantID, fished_or_hired_3months_i1, leased_now_i31, 
         self_now_i31, hired_now_i31, forpay_now_i31, enough_fish_i32, starts_with(c("quantity", "limits", "change", "shock")))
```

Replacing values of fisher variables with NA for non-fishers (non-fishers = `fished_or_hired_3months == 0`)

```{r}

fishing_behavior_f <- fishing_behavior_f %>% 
  mutate_at(vars(hired_now_i31, self_now_i31, forpay_now_i31, leased_now_i31, 
                 starts_with(c("quantity_", "limits", "change", "shock"))), ~replace(.,fished_or_hired_3months_i1 == 0, NA))
            
```


In some time points, the fishing access questionnaire includes likert scale questions about perecpetions/preferences related to fishing. Currently excluding these because it's not clear how I will use them, but I may bring them in later.

Next step is to combine the fisher and non-fisher questionnaires in a way that the overlapping questions merge, and the remaining questions append.




**ORGANIZE AND EXPLORE RELEVANT MORBIDITY QUESTIONS**

```{r}
mh <- morbidity %>% 
  mutate(health_l1 = L1_HealthStatus,
         pain_l1 = L2_PainPastThirtyDays,
         social_limit_l121 = L121_HealthLimitSocialActivities_FreqID, 
         anxious_l122 = L122_NervousAnxious_FreqID,
         calm_l123 = L123_CalmPeaceful_FreqID,
         depressed_l124 = L124_Depressed_FreqID,
         happy_l125 = L125_Happy_FreqID,
         depressed_extra_l126 = L126_VeryDepressed_FreqID,
         energy_full_l127 = L127_FullLifeEnergy_FreqID, 
         energy_none_l128 = L128_WithoutEngery_FreqID,
         tired_l129 = L129_Tired_FreqID,
         energy_enough_l1210 = L1210_EnoughEnergy_FreqID,
         weighed_down_health_l1211 = L1211_WeighedDownByHealth_FreqID,
         discouraged_health_l1212 = L1212_DiscouragedByHealth_FreqID,
         despair_health_l1213 = L1213_DespairOverHealth_FreqID, 
         afraid_health_l1214 = L1214_AfraidBecauseHealth_FreqID, 
         reasoning_diffiuclty_l1215 = L1215_DifficultyReasoningDecisions_FreqID, 
         forgetting_l1216 = L1216_ForgotThings_FreqID, 
         attention_l1217 = L1217_TroubleKeepingAttention_FreqID,
         concentration_l1218 = L1218_DifficultyConcentrationThinking_FreqID,
         lifequality_l14 = L14_QualityOfLife) %>% 
  select(TimePoint, ParticipantID, (starts_with(c("health", "pain", "social", "anxious", "calm", "depressed", "happy", "energy", "tired", "weighed", "discouraged", "despair", "afraid", "reasoning", "forgetting", "attention", "concentration", "life" ))))

```

